\ripple: Differentiable and Hardware-Accelerated Waveforms for Gravitational Wave Data Analysis

### 2023-01-24 src/tex/ms.tex
bunch of updates

% Abstract with filler text
\begin{abstract}
    % Since their discovery in 2015, gravitational waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
    % As more detections continue, data analysis tasks are becoming increasingly computationally demanding.
    % \kw{I would cut everything before out.}
    We propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of analysis tasks throughout gravitational wave (GW) science.
    Firstly, we demonstrate that complete waveforms which cover the inspiral, merger, and ringdown of binary black holes (i.e.~IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
    Moreover, \jax allows for GPU-accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
    We then focus on three applications where efficient and differentiable waveforms are essential.
    Firstly, we demonstrate how gradient descent can be used to optimize the $\sim 200$ coefficients that are used to calibrate the waveform model. 
    In particular, we demonstrate that the typical \textit{match} with numerical relatively waveforms can be improved by more than 10\% without any additional overhead.
    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on a CPU) with no loss in accuracy.
    This increased speed makes population forecasting substantially simpler.
    % , as discussed extensively in \citet{Iacovelli:2022bbs, Iacovelli:2022mbg}.
    Finally, we show that gradient-based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to traditional Monte Carlo methods.
    Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
    Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are implemented.

@@

The complexity begins even before data taking, since GW searches using the matched-filtering technique~\citep{Owen:1998dk, Owen:1995tm} require the generation of large banks of template waveforms.
% \sout{The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. }
Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
For binary black holes with non-aligned spins, this requires a Markov Chain Monte Carlo (MCMC) on an 15 dimensional parameter space.
More general examples can lead to a significant increase in dimension.\te{add citations}
% In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
% \kw{This is not quite true. You can always marginalize over more parameters if you want, such as time of coalescence and phase. I would just mention the typical case of 15 parameters GR PE instead.}
% \AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
Beyond this simple scenario, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio, Yunes:2016jcc, Arun:2006yw, Agathos:2013upa}.
% \AZ{Add PPE and maybe TIGER refs..?}
Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, LIGOScientific:2021psn, Wong:2022flg}.
\kw{Need more reference and some rewording}\te{added refs, happy?}
% All of these tasks require significant computation.
Overall, analyzing GWs therefore requires significant computation.
% \kw{This could be triggering to some people, since population analysis could be quite cheap}
In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleviating this computational demand.

@@

\AZ{Add refs here:}\te{Added a bunch, let me know if more are needed!}
There are a variety of gravitational waveforms currently used in analysis pipelines.
They are generally structured into different families, the most common of which are: the effective-one-body (EOB)~\citep{Damour:2008yg,Buonanno:2005xu, Buonanno:2000ef, Buonanno:1998gg, Damour:2000we}, the phenomenological inspiral-merger-ringerdown (IMRPhenom), and numerical relativity surrogate (NRsurrogate)~\citep{Varma:2018mmi,Varma:2019csw}.
%Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
Of these, the IMRPhenom family serves as a natural starting point for an AD implementation in \jax.
Models like the non-precessing IMRPhenomD~\citep{Khan:2015jqa} model studied here and the precessing, higher-mode model IMRPhenomXPHM~\citep{Pratten:2020fqn, Pratten:2020ceb} are written in the frequency domain using closed-form expressions.
This makes a
% \sout{In particular, their closed form expressions and fixed size frequency grids make a }
\jax implementation that complies with the constraints of JIT compilation simple.
NRsurrogate models, which interpolate directly over waveforms produced by numerical relativity simulations, are in principle also straightfoward to implement in \jax.
EOB waveforms on the other hand are produced by evolving the dynamics of an effective one body Hamiltonian, and are therefore more difficult to implement in \jax.
For EOB waveforms, frequency-domain reduced-order models may be a convenient target~\citep[e.g.][]{Cotesta:2020qhw}.
% to produce a waveform
%require dynamic frequency grids and are therefore tricky to implement in \jax.
%\AZ{What does this mean? I don't understand why EOB needs dynamic frequency grids. We should clarify, and we can also mention if \jax is well-suited to frequency-domain ROM implementations of EOB, and reference those.} 
%In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
%\AZ{rephrase this in a more positive fashion:}
%NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cumbersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}

@@

A variety of waveform families have been developed to accurately model the GW emission from COs~\citep{Schmidtreview}. 
When the COs are relatively well separated, the dynamics of the system can be well approximated by a post-Newtonian expansion.
However, close to merger, numerical relatively (NR) simulations are required to accurately model the binary.
Unfortunately, these numerical simulations are computationally expensive and cannot be run in conjunction with data analysis.
Approximate, phenomenological waveforms have therefore been constructed to enable relatively fast waveform generation at sufficient accuracy.

As mentioned in the previous section, the three major waveform families that have been developed to date are:  EOB,  NRsurrogate, and IMRPhenom.
Note that unlike NRsurrogate, both the EOB and IMRPhenom waveforms must be calibrated to numerical relativity waveforms.
EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate, whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
IMRPhenom waveforms are therefore ideally-suited for AD, especially using \jax. 
In this paper we focus on the aligned-spin, circular-orbit model IMRPhenomD~\citep{Husa:2015iqa, Khan:2015jqa}.

@@

To demonstrate our implementation of IMRPhenomD is faithful to the \lalsuite implementation, we start by defining the noise weighted inner product:
\begin{equation}
    \label{eq:inner_prod}
    \left(h_1|h_2\right) \equiv 4 \, \mathrm{Re} \int^{\infty}_{0} \mathrm{d} f \, \frac{ h^*_1(f) h_2(f)}{S_n(f)}\, ,
\end{equation}
where $S_n$ is the (one-sided) noise power spectral density (PSD) and \AZ{$h_1$ and $h_2$ are the frequency domain waveforms which are to be compared.}
\sout{ evaluated at the intrinsic parameter point $\theta_1$.}
We can then normalize the inner product through
\begin{equation}
    \left[h_1|h_2\right] = \frac{\left(h_1|h_2\right)}{\sqrt{\left(h_1|h_1\right)\left(h_2|h_2\right)}}\, .
\end{equation}
Now we are ready to define the \textit{match} which is given by 
\begin{equation}
    \mathrm{m}(h_1|h_2) \equiv \max_{\Delta t_c,\, \Delta \phi_c} \left[h_1 \, \middle| \, h_2 \right]\,
\end{equation}
where \AZ{$\Delta t_c$ and $\Delta \phi_c$ are, respectively, the differences in time and phase of coalescence between the two waveforms.}

Since the match is a measure of the difference between two waveforms, we can use it to demonstrate that the implementation of IMRPhenomD in \ripple accurately matches the \lalsuite implementation.
\AZ{Say what PSD was used and reference it.}
This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_1|h_2)$ across the entire parameter space.\footnote{
    Specifically, we use $10^4$ points \AZ{varying component masses $m_1$, $m_2$ and spin parameters $\chi_1$, $\chi_2$} in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$. In addition, we evaulated the waveforms on a frequency grid from $32\,$Hz to $1024\,$Hz with frequency spacing $\Delta f = 0.0125 \,$Hz.
}
\AZ{Here, $h_1$ corresponds to the \ripple waveform implementation and $h_2$ is the \lalsuite implementation, evaluated at the same point in parameter space.} 
From Fig.~\ref{fig:match}, it is clear that the \ripple waveform matches with the \lalsuite waveform close to numerical precision across the entire parameter space.
\AZ{strictly $=0$? That seems impossible. Fmla has wrong labels for $h$'s}
In particular, the gray points at low masses are where $1-\mathrm{m}(h_1|h2)=0$ causing points to not appear in the log scale colorbar.
% In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
\begin{figure}[t]
    \script{random_matches.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/random_matches.pdf}
        \caption{
            Match between the \ripple and \lalsuite implementations of the IMRPhenomD waveform as a function of total mass and effective spin. 
            % The lower density of points at low masses is caused by the match being exactly 1.0 and therefore not appearing in the log scale. 
            The gray points are where the match is exactly 1.0 and therefore would not appear in the log scale.
            It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
            \dfm{I think this figure could use some work. Maybe the points could be smaller to make them easier to resolve. And it could be interesting to clip the mismatch to something like $10^{-16}$ so that we don't lose the points with zero mismatch. Right now it looks less impressive!}
        }
        \label{fig:match}
    \end{centering}
\end{figure}

Note that in General Relativity (GR) the total mass $M = m_1 + m_2$ of \AZ{a binary black hole simply serves as an overall scale for the system, so that the frequency evolution can be trivially rescaled.
The total mass only impacts the match because the chosen PSD and frequency limits fix a reference scale.}
\sout{the system is simply an overall scaling of the waveforms} 
\sout{frequency evolution and is therefore not a} \sout{truly independent parameter.}
If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and \AZ{rescale the frequency grid to units of $Mf$}, \sout{correctly,} all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
\AZ{Revisit this after defining PSD above:} 
This figure instead illustrates a more realistic PSD where at low masses the waveform signal-to-noise is dominated by the inspiral but at high masses it is dominated by the merger.

There remains some slight deviation between the two waveform implementations at high total mass. 
This is partly due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \jax. \dfm{What dimension of interpolation do you need? Maybe this is something we could implement...}
Instead, we initially use \texttt{scipy}'s cubic interpolator to create a fine grid of $5\times10^5$ values, which we then linearly interpolate during waveform generation.
Unfortunately, we cannot make the initial cubic interpolation arbitrarily fine as this would add additional computational overhead when loading the data during waveform evaluation.
% The remaining error in the waveform at high masses is therefore primarily down to the coarseness in this interpolation.
Note however, that the differences are well below the accuracy requirements of the waveforms and will have no noticeable for realistic data analysis tasks.

For maximum utility, a waveform needs to be fast to evaluate.
Fortunately, the IMRPhenom waveforms are constructed from simple closed-form expressions which are computationally efficient.
Since the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the evaluation of the waveform itself is faster than \ripple. \dfm{I'd soften this statement a little bit. In my own work, I often find that JAX implementations are just as fast or faster than C implementations, so I wouldn't state it that it's obvious that lalsuite should be faster! Where are these benchmarks implemented? Would it be easy for me to try them out?}
Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 2-5\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 5-8\,\mathrm{ms}$ for \ripple.\footnote{
    For this benchmark we used $16\,\mathrm{Hz}$ and $1024\,\mathrm{Hz}$ for the lowest and highest frequencies respectively. 
    In addition, we used a frequency spacing of $0.0125\,\mathrm{Hz}$. 
    We performed this benchmark by evaluating the waveform $5000$ times and taking the average evaluation time.
}
Although slower on a CPU benchmark, a key advantage of using \jax is its native ability to run on a GPU.
Waveform calls can therefore be evaluated in parallel, massively improving performance.
Performing the same benchmark as above on a Google Colab notebook with a freely-provided NVIDIA Tesla T4 GPU (far from the current state-of-the-art), we find that on average waveform evaluations take $\sim \sout{0.00005} 5 \times 10^{-5}\mathrm{ms}$, nearly five orders of magnitude faster.
\kw{We should add some more detail related to this comparison. We should be clear that each individual waveform sample evaluation doesn't get us the performance by putting it on GPU (Jitting helps with that, not the GPU part). The GPU comes into play by allowing parallel evaluation of waveform sample, so for a tiny amount of time/frequency samples, we should expect GPU to lose to a CPU, but on any reasonable amount of time/frequency samples, GPU win a lot.}
\amc{mention \texttt{vmap} and/or \texttt{pmap} -- super trivial to parallelize! More work in C.}
Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.

The primary argument of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
First, the accuracy of derivatives from AD are significantly more stable than finite-difference methods.
In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
On the other hand, AD produces machine-precision derivatives with no tuning.
Second, AD scales favorably with the dimensionality of the function.
In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
For reverse-mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimension.\footnote{
    Note that although the number of function calls is small, reverse-mode AD does add memory overhead.
    We've not found this to be limiting in any of the situations tested so far.
}
\kw{Maybe a bit more background on AD, the way we phrase it here sounds a bit too magical too me now.}
Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
However, this might change for waveforms in beyond-GR models, \AZ{waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
In these cases} many more parameters \AZ{can be} added.

@@

Having an accurate waveform model is essential for many data analysis tasks in
GW, such as searching for signals and estimating source parameters given data.
While waveforms generated using numerical relativity simulations are \sout{commonly
regarded as the most accurate,} 
\AZ{in principle the highest fidelity signal model,}
they are too computationally expensive to be used
in any practical data analysis tasks. To create waveform models that can be used
in data analysis, there are a number of groups creating ``approximants'' of the
full numerical relativity waveforms, which are essentially simpler ansatzes that
\sout{are only approximations but} can be evaluated much faster \AZ{and in regions not
covered by numerical simulations, such as extremes of parameter space or earlier 
phases of the binary inspiral}. 
\amc{redundant -- remove or merge with earlier material.}

@@

\begin{figure}[t]
	\script{loss_compare.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/loss_compare.pdf}
		\caption{
			Relative error between the numerical relatively and IMRPhenomD waveform amplitudes. 
            The blue line shows the relative error of the IMRPhenomD model with the original model coefficients where as the orange line uses the IMRPhenomD model after optimization using gradient descent.
            Interestingly, the error is reduced across the frequency range, demonstrating that high dimensional optimization is useful for all parts of the waveform.
		}
		\label{fig:loss_compare}
	\end{centering}
\end{figure}

>>>>

% Abstract with filler text
\begin{abstract}
    % Since their discovery in 2015, gravitational waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
    % As more detections continue, data analysis tasks are becoming increasingly computationally demanding.
    % \kw{I would cut everything before out.}
    We propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of analysis tasks throughout gravitational wave (GW) science.
    Firstly, we demonstrate that complete waveforms which cover the inspiral, merger, and ringdown of binary black holes (i.e.~IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \texttt{C}.
    Moreover, \jax allows for GPU-accelerated waveform calls which can be over an order of magnitude faster than serial evaluation on a CPU.
    We then focus on three applications where efficient and differentiable waveforms are essential.
    Firstly, we demonstrate how gradient descent can be used to optimize the $\sim 200$ coefficients that are used to calibrate the waveform model. 
    In particular, we demonstrate that the typical \textit{match} with numerical relativity waveforms can be improved by more than 10\% without any additional overhead.
    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on a CPU) with no loss in accuracy.
    This increased speed makes population forecasting substantially simpler.
    % , as discussed extensively in \citet{Iacovelli:2022bbs, Iacovelli:2022mbg}.
    Finally, we show that gradient-based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to traditional Monte Carlo methods.
    Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
    Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are implemented.

@@

The complexity begins even before data taking, since GW searches using the matched-filtering technique~\citep{Owen:1998dk, Owen:1995tm} require the generation of large banks of template waveforms.
% \sout{The complexity begins before data gathering where one is required to generate banks of template waveforms which will be used for a matched-filter search~\citep{Owen:1998dk, Owen:1995tm}. }
Once potential candidates are found, parameter estimation (PE) is performed to extract the detailed source properties of each event~\citep{Christensen:2022bxb, 2020MNRAS.493.3132S, Ashton:2018jfp, Romero-Shaw:2020owr, Veitch:2014wba, Biwer:2018osg}.
For binary black holes with non-aligned spins, this requires a Markov Chain Monte Carlo (MCMC) on a 15 dimensional parameter space.
More general binary inspirals, such as those involving neutron stars, can lead to a significant increase in dimension.\te{add citations}
% In the most minimal example of non-precessing binaries, this requires a Markov Chain Monte Carlo (MCMC) on an 11 dimensional parameter space for each event.
% \kw{This is not quite true. You can always marginalize over more parameters if you want, such as time of coalescence and phase. I would just mention the typical case of 15 parameters GR PE instead.}
% \AZ{Maybe modify; some other points are precessing PE requires 15 params, tidal prec 17 params, plus calibration marge and EoS inferences can incorporate even more}
Beyond these simple scenarios, more complex waveform models with additional parameters may be used to test for deviations from General Relativity~\citep{Krishnendu:2021fga, LIGOScientific:2020tif, LIGOScientific:2016lio, Yunes:2016jcc, Arun:2006yw, Agathos:2013upa}.
% \AZ{Add PPE and maybe TIGER refs..?}
Finally, using the results of PE, population synthesis models constrain the progenitors systems from which the black holes we see merging today began their journey~\citep{LIGOScientific:2020kqk, LIGOScientific:2021psn, Wong:2022flg}.
\kw{Need more reference and some rewording}\te{added refs, happy?}
% All of these tasks require significant computation.
Overall, GW data analysis therefore requires significant computation.
% \kw{This could be triggering to some people, since population analysis could be quite cheap}
In this paper, we will argue that differentiable waveforms (and more generally differentiable pipelines) can play a significant role in alleviating this computational demand.

@@

\AZ{Add refs here:}\te{Added a bunch, let me know if more are needed!}
There are a variety of gravitational waveforms currently used in analysis pipelines.
They are generally structured into different families, the most common of which are: the effective-one-body (EOB)~\citep{Damour:2008yg,Buonanno:2005xu, Buonanno:2000ef, Buonanno:1998gg, Damour:2000we}, the phenomenological inspiral-merger-ringerdown (IMRPhenom), and numerical relativity surrogate (NRsurrogate)~\citep{Varma:2018mmi,Varma:2019csw}.
%Of these, the IMRPhenom family is most well suited for an AD implementation in \jax.
Of these, the IMRPhenom family serves as a natural starting point for an AD implementation in \jax.
Models like the non-precessing IMRPhenomD~\citep{Khan:2015jqa} model studied here and the precessing, higher-mode model IMRPhenomXPHM~\citep{Pratten:2020fqn, Pratten:2020ceb} are written in the frequency domain using closed-form expressions.
This makes a
% \sout{In particular, their closed form expressions and fixed size frequency grids make a }
\jax implementation that complies with the constraints of JIT compilation simple.
NRsurrogate models, which interpolate directly over waveforms produced by numerical relativity (NR) simulations, are in principle also straightfoward to implement in \jax.
EOB waveforms on the other hand are produced by evolving the dynamics of an effective one body Hamiltonian, and are therefore more difficult to implement in \jax.
For EOB waveforms, frequency-domain reduced-order models may be a convenient target~\citep[e.g.][]{Cotesta:2020qhw}.
% to produce a waveform
%require dynamic frequency grids and are therefore tricky to implement in \jax.
%\AZ{What does this mean? I don't understand why EOB needs dynamic frequency grids. We should clarify, and we can also mention if \jax is well-suited to frequency-domain ROM implementations of EOB, and reference those.} 
%In addition, EOB waveforms use hypergeometric functions which have limited support in \jax.
%\AZ{rephrase this in a more positive fashion:}
%NRsurrogate waveforms have no fundamental issue but require the user to download relatively large data files, making their use cumbersome in practice.\te{Max/Kaze, please add to/change this if you think its incorrect}

@@

A variety of waveform families have been developed to accurately model the GW emission from COs~\citep{Schmidtreview}. 
When the COs are relatively well separated, the dynamics of the system can be well approximated by a post-Newtonian expansion.
However, close to merger, NR simulations are required to accurately model the binary.
Unfortunately, these numerical simulations are computationally expensive and cannot be run in conjunction with data analysis.
Approximate, phenomenological waveforms have therefore been constructed to enable relatively fast waveform generation at sufficient accuracy.

As mentioned in the previous section, the three major waveform families that have been developed to date are:  EOB,  NRsurrogate, and IMRPhenom.
Note that unlike NRsurrogate, both the EOB and IMRPhenom waveforms must be calibrated to NR waveforms.
EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate, whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
IMRPhenom waveforms are therefore ideally-suited for AD, especially using \jax. 
In this paper we focus on the aligned-spin, circular-orbit model IMRPhenomD~\citep{Husa:2015iqa, Khan:2015jqa}.

@@

To demonstrate our implementation of IMRPhenomD is faithful to the \lalsuite implementation, we start by defining the noise weighted inner product:
\begin{equation}
    \label{eq:inner_prod}
    \left(h_1|h_2\right) \equiv 4 \, \mathrm{Re} \int^{\infty}_{0} \mathrm{d} f \, \frac{ h^*_1(f) h_2(f)}{S_n(f)}\, ,
\end{equation}
where $S_n$ is the (one-sided) noise power spectral density (PSD) and $h_1$ and $h_2$ are the frequency domain waveforms which are to be compared.
% \sout{ evaluated at the intrinsic parameter point $\theta_1$.}
We can then normalize the inner product through
\begin{equation}
    \left[h_1|h_2\right] = \frac{\left(h_1|h_2\right)}{\sqrt{\left(h_1|h_1\right)\left(h_2|h_2\right)}}\, .
\end{equation}
Now we are ready to define the \textit{match} which is given by 
\begin{equation}
    \mathrm{m}(h_1|h_2) \equiv \max_{\Delta t_c,\, \Delta \phi_c} \left[h_1 \, \middle| \, h_2 \right]\,
\end{equation}
where $\Delta t_c$ and $\Delta \phi_c$ are, respectively, the differences in time and phase of coalescence between the two waveforms.

Since the match is a measure of the difference between two waveforms, we can use it to demonstrate that the implementation of IMRPhenomD in \ripple accurately matches the \lalsuite implementation. For this comparison, we use the $S_n$  presented in GWTC-2~\citep{LIGOScientific:2020ibl} from the Livingston detector (the most sensitive of the detectors).\footnote{
    \url{https://dcc.ligo.org/LIGO-P2000251/public}.
    }
This is shown in Fig.~\ref{fig:match}, where we have calculated the $\mathrm{m}(h_1|h_2)$ across the entire parameter space.\footnote{
    Specifically, we use $10^4$ points varying component masses $m_1$, $m_2$ and spin parameters $\chi_1$, $\chi_2$ in the ranges $m_{1,2} = (1,100)\,M_{\odot}$ and $\chi_{1,2} = (-1,1)$. In addition, we evaulated the waveforms on a frequency grid from $32\,$Hz to $1024\,$Hz with frequency spacing $\Delta f = 0.0125 \,$Hz.
}
Here, $h_1$ corresponds to the \ripple waveform implementation and $h_2$ is the \lalsuite implementation, evaluated at the same point in parameter space.
From Fig.~\ref{fig:match}, it is clear that the \ripple waveform matches with the \lalsuite waveform close to numerical precision across the entire parameter space.
\AZ{strictly $=0$? That seems impossible. Fmla has wrong labels for $h$'s}
In particular, the gray points at low masses are where $1-\mathrm{m}(h_1|h_2)=0$ causing points to not appear in the log scale colorbar.
% In particular, the blue points indicate parameter points where the $1-\mathrm{m}(h_{\theta_1}|h^{*}_{\theta_1})=0$ and would therefore not appear in the log scale colorbar.
\begin{figure}[t]
    \script{random_matches.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/random_matches.pdf}
        \caption{
            Match between the \ripple and \lalsuite implementations of the IMRPhenomD waveform as a function of total mass and effective spin. 
            % The lower density of points at low masses is caused by the match being exactly 1.0 and therefore not appearing in the log scale. 
            The gray points are where the match is exactly 1.0 and therefore would not appear in the log scale.
            It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
            \te{Current plot looking better?}
            \dfm{I think this figure could use some work. Maybe the points could be smaller to make them easier to resolve. And it could be interesting to clip the mismatch to something like $10^{-16}$ so that we don't lose the points with zero mismatch. Right now it looks less impressive!}
        }
        \label{fig:match}
    \end{centering}
\end{figure}

Note that in General Relativity (GR) the total mass $M = m_1 + m_2$ of a binary black hole simply serves as an overall scale for the system, so that the frequency evolution can be trivially rescaled.
The total mass only impacts the match because the chosen PSD and frequency limits fix a reference scale.
% \sout{the system is simply an overall scaling of the waveforms} 
% \sout{frequency evolution and is therefore not a} \sout{truly independent parameter.}
If we instead use a flat PSD in Eq.~\ref{eq:inner_prod} and rescale the frequency grid to units of $Mf$, all dependence with $M$ seen in Fig.~\ref{fig:match} vanishes. 
% \AZ{Revisit this after defining PSD above:} 
This figure instead illustrates a more realistic PSD where at low masses the waveform signal-to-noise is dominated by the inspiral but at high masses it is dominated by the merger.

There remains some slight deviation between the two waveform implementations at high total mass. 
This is partly due to the fact that cubic interpolators, which are used within IMRPhenomD to calculate the ringdown and damping frequencies, are not currently supported in \jax. \dfm{What dimension of interpolation do you need? Maybe this is something we could implement...}
\te{Its just one dimension, so is pretty easy. I tried the one from the cosmoJAX people and found that the match did slightly increase but it was a decent chunk slower (maybe a factor of 2 or so). So I just left it out. Would be happy to code up something and put it in if you think its necessary!}
Instead, we initially use \texttt{scipy}'s cubic interpolator to create a fine grid of $5\times10^5$ values, which we then linearly interpolate during waveform generation.
Unfortunately, we cannot make the initial cubic interpolation arbitrarily fine as this would add additional computational overhead when loading the data during waveform evaluation.
% The remaining error in the waveform at high masses is therefore primarily down to the coarseness in this interpolation.
Note however, that the differences are well below the accuracy requirements of the waveforms and will have no noticeable effect for realistic data analysis tasks.

For maximum utility, a waveform needs to be fast to evaluate.
Fortunately, the IMRPhenom waveforms are constructed from simple closed-form expressions which are computationally efficient.
Even though the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the serial evaluation of the waveform in \ripple is comparably fast. 
% \dfm{I'd soften this statement a little bit. In my own work, I often find that JAX implementations are just as fast or faster than C implementations, so I wouldn't state it that it's obvious that lalsuite should be faster! Where are these benchmarks implemented? Would it be easy for me to try them out?}
Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 0.5\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 0.4\,\mathrm{ms}$ for \ripple.\footnote{
    For this benchmark we used $24\,\mathrm{Hz}$ and $512\,\mathrm{Hz}$ for the lowest and highest frequencies respectively. 
    In addition, we used a frequency spacing of $0.2\,\mathrm{Hz}$. 
    We performed this benchmark by evaluating the waveform $10^{4}$ times and taking the average evaluation time.
}
Although similar on a CPU benchmark, \jax has a few key advantages. 
First, its ability to JIT compile allows for significant performace gains (the above benchmark is already JIT compiled).
Second, automatic vectorization can be achieved using the \texttt{vmap} function.
Using \texttt{vmap} and performing the same benchmark as described above reduces the average evaluation speed to $\sim 0.14\,\mathrm{ms}$.
Finally, \jax can natively on a GPU which allows for highly parallellized waveform evaluations.

% Waveform calls can therefore be evaluated in parallel, massively improving performance.

Again, performing the same benchmark as above on a NVIDIA Quadro 6000, we find that on average waveform evaluations take $\sim 0.02$, over an of magnitude faster than serial CPU evaluation.
% \te{Check and modify}
% \kw{We should add some more detail related to this comparison. We should be clear that each individual waveform sample evaluation doesn't get us the performance by putting it on GPU (Jitting helps with that, not the GPU part). The GPU comes into play by allowing parallel evaluation of waveform sample, so for a tiny amount of time/frequency samples, we should expect GPU to lose to a CPU, but on any reasonable amount of time/frequency samples, GPU win a lot.}
% \amc{mention \texttt{vmap} and/or \texttt{pmap} -- super trivial to parallelize! More work in C.}
Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.

One of the primary argument of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
First, the accuracy of derivatives from AD are significantly more stable than finite-difference methods.
In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
On the other hand, AD produces machine-precision derivatives with no tuning.
Second, AD scales favorably with the dimensionality of the function.
In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
For reverse-mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimension.\footnote{
    Note that although the number of function calls is small, reverse-mode AD does add memory overhead.
    We've not found this to be limiting in any of the situations tested so far.
}
\kw{Maybe a bit more background on AD, the way we phrase it here sounds a bit too magical too me now.}
Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
However, this might change for waveforms in beyond-GR models, waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
In these cases many more parameters can be added.

@@

Having an accurate waveform model is essential for many data analysis tasks in
GW, such as searching for signals and estimating source parameters given data.
While waveforms generated using NR simulations are \sout{commonly
regarded as the most accurate,} 
\AZ{in principle the highest fidelity signal model,}
they are too computationally expensive to be used
in any practical data analysis tasks. To create waveform models that can be used
in data analysis, there are a number of groups creating ``approximants'' of the
full NR waveforms, which are essentially simpler ansatzes that
\sout{are only approximations but} can be evaluated much faster \AZ{and in regions not
covered by numerical simulations, such as extremes of parameter space or earlier 
phases of the binary inspiral}. 
\amc{redundant -- remove or merge with earlier material.}

@@

\begin{figure}[t]
	\script{loss_compare.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/loss_compare.pdf}
		\caption{
			Relative error between the numerical relativity and IMRPhenomD waveform amplitudes. 
            The blue line shows the relative error of the IMRPhenomD model with the original model coefficients where as the orange line uses the IMRPhenomD model after optimization using gradient descent.
            Interestingly, the error is reduced across the frequency range, demonstrating that high dimensional optimization is useful for all parts of the waveform.
		}
		\label{fig:loss_compare}
	\end{centering}
\end{figure}
